Nodal Discontinuous Galerkin Method for the Euler Equations with Curvilinear Coordinates and Nuclear Equation of State for Astrophysics Simulations

### 2016-08-03 Documents/Euler/dgEulerCurvilinear.tex
Starting to add gravitational sources to equations

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage{times}
\usepackage{url}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{yhmath}
\usepackage{natbib}
\bibliographystyle{apj}
\usepackage{../definitions}
\hypersetup{
  bookmarksnumbered = true,
  bookmarksopen=false,
  pdfborder=0 0 0,         % make all links invisible, so the pdf looks good when printed
  pdffitwindow=true,      % window fit to page when opened
  pdfnewwindow=true, % links in new window
  colorlinks=true,           % false: boxed links; true: colored links
  linkcolor=blue,            % color of internal links
  citecolor=magenta,    % color of links to bibliography
  filecolor=magenta,     % color of file links
  urlcolor=cyan              % color of external links
}

@@

$\vect{r}(\vect{x})\in\mathbb{R}^{3}$, so that the displacement vector is $d\vect{r}=dx^{i}\,\vect{e}_{i}$, where the coordinate basis vectors are
\begin{equation}
  \vect{e}_{i}=\pderiv{\vect{r}}{x^{i}}
\end{equation}
$ds_{\vect{x}}^{2}=d\vect{r}\cdot d\vect{r}$
\begin{equation}
  ds_{\vect{x}}^{2}=\gamma_{ij}\,dx^{i}\,dx^{j},
\end{equation}
where the coordinate basis metric tensor is defined as $\gamma_{ij}=\vect{e}_{i}\cdot\vect{e}_{j}$.

For general curvilinear spatial coordinates, the Euler equations for gas dynamics are \citep{landauLifshitz_1959}
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\rho\,v^{i}\,\Big)=0,
\end{equation}
\begin{equation}
  \pderiv{M_{j}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,P^{i}_{~j}\,\Big)=\f{1}{2}\,P^{ik}\,\pderiv{\gamma_{ik}}{x^{j}},
\end{equation}
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\big(E+p\big)\,v^{i}\,\Big)=0,
\end{equation}
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,n_{e}\,v^{i}\,\Big)=0,
\end{equation}
where $\rho$ is the (baryon) mass density, $M^{i}=\rho\,v^{i}$ is the momentum density, $P^{ij}=\rho\,v^{i}\,v^{j}+p\,\gamma^{ij}$ the stress tensor, $E=e+\f{1}{2}\,\rho\,v_{i}\,v^{i}$ the total (internal plus kinetic) energy density, and $n_{e}$ the electron density.  
The pressure is related to the density, internal energy density, and electron density by an equation of state (EoS); i.e., $p=p(\rho,e,n_{e})$.  
(We also define the specific internal energy by $\epsilon=e/\rho$.)

@@

The mass conservation equation
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\rho\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\rho\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\rho\,\bar{v}^{3}\,\Big)=0
\end{equation}
Components of the momentum equation
\begin{equation}
  \pderiv{\bar{M}^{i}}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\bar{P}^{1i}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\bar{P}^{2i}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\bar{P}^{3i}\,\Big)
  =\bar{S}^{i},
\end{equation}
where the ``geometry sources'' are given by
\small
\begin{equation}
  \bar{S}^{1}
  =\bar{P}^{22}\f{1}{a}\pderiv{a}{x^{1}}+\bar{P}^{33}\f{1}{b}\pderiv{b}{x^{1}}, \quad
  \bar{S}^{2}
  =\bar{P}^{33}\f{1}{ac}\pderiv{c}{x^{2}}-\bar{P}^{12}\f{1}{a}\pderiv{a}{x^{1}}, \quad
  \bar{S}^{3}
  =-\bar{P}^{13}\f{1}{b}\pderiv{b}{x^{1}}-\bar{P}^{23}\f{1}{ac}\pderiv{c}{x^{2}}.  
\end{equation}
\normalsize
Energy equation
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\big(E+p\big)\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\big(E+p\big)\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\big(E+p\big)\,\bar{v}^{3}\,\Big)=0.  
\end{equation}
The electron conservation equation
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,n_{e}\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,n_{e}\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,n_{e}\,\bar{v}^{3}\,\Big)=0
\end{equation}
By defining $\bU=\big(\,\rho,\,\rho\,\bar{v}^{1},\,\rho\,\bar{v}^{2},\,\rho\,\bar{v}^{3},\,E,\,n_{e}\,\big)^{T}$, $\vect{S}(\bU)=\big(\,0,\,\bar{S}^{1},\,\bar{S}^{2},\,\bar{S}^{3}\,,0,\,0\,\big)^{T}$, and
\begin{equation}
  \vect{F}^{i}(\bU)
  =\big(\,\rho\,\bar{v}^{i},\,\rho\,\bar{v}^{1}\,\bar{v}^{i}+p\,\delta^{1i},\,\rho\,\bar{v}^{2}\,\bar{v}^{i}+p\,\delta^{2i},\,\rho\,\bar{v}^{3}\,\bar{v}^{i}+p\,\delta^{3i},\,(E+p)\,\bar{v}^{i},\,n_{e}\,\bar{v}^{i}\,\big)^{T},
\end{equation}
we can write the system of equations in compact form
\begin{equation}
  \pderiv{\bU}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\vect{F}^{1}(\bU)\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\vect{F}^{2}(\bU)\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\vect{F}^{3}(\bU)\,\Big)
  =\vect{S}(\bU).
  \label{eq:extendedEulerCompact}
\end{equation}

@@

\bibliography{./references/references.bib}

>>>>

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{mathrsfs}
\usepackage{cite}
\usepackage{times}
\usepackage{url}
\usepackage{hyperref}
\usepackage{lineno}
\usepackage{yhmath}
\usepackage{natbib}
\usepackage{../definitions}
\hypersetup{
  bookmarksnumbered = true,
  bookmarksopen=false,
  pdfborder=0 0 0,         % make all links invisible, so the pdf looks good when printed
  pdffitwindow=true,      % window fit to page when opened
  pdfnewwindow=true, % links in new window
  colorlinks=true,           % false: boxed links; true: colored links
  linkcolor=blue,            % color of internal links
  citecolor=magenta,    % color of links to bibliography
  filecolor=magenta,     % color of file links
  urlcolor=cyan              % color of external links
}

@@

We consider evolution in three-dimensional position space and denote the position vector with for a point with coordinates $\bx=\{x^{1},x^{2},x^{3}\}$ with $\vect{r}(\vect{x})\in\mathbb{R}^{3}$, so that the displacement vector is $d\vect{r}=dx^{i}\,\vect{e}_{i}$ and the coordinate basis vectors are
\begin{equation}
  \vect{e}_{i}=\pderiv{\vect{r}}{x^{i}}.  
\end{equation}
The squared infinitesimal line-element is
\begin{equation}
  ds_{\vect{x}}^{2}=d\vect{r}\cdot d\vect{r}=\gamma_{ij}\,dx^{i}\,dx^{j},
\end{equation}
where the coordinate basis metric tensor is defined as $\gamma_{ij}=\vect{e}_{i}\cdot\vect{e}_{j}$.

For general curvilinear spatial coordinates, the Euler equations for gas dynamics are \citep{leveque_2002}
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\rho\,v^{i}\,\Big)=0,
\end{equation}
\begin{equation}
  \pderiv{M_{j}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,P^{i}_{~j}\,\Big)
  =\f{1}{2}\,P^{ik}\,\pderiv{\gamma_{ik}}{x^{j}}-\rho\,\pderiv{\Phi}{x^{j}},
\end{equation}
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,\big(E+p\big)\,v^{i}\,\Big)
  =-M^{i}\,\pderiv{\Phi}{x^{i}},
\end{equation}
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{\sqrt{\gamma}}\pderiv{}{x^{i}}\Big(\,\sqrt{\gamma}\,n_{e}\,v^{i}\,\Big)=0,
\end{equation}
where $\rho$ is the (baryon) mass density, $M^{i}=\rho\,v^{i}$ is the momentum density, $P^{ij}=\rho\,v^{i}\,v^{j}+p\,\gamma^{ij}$ the stress tensor, $E=e+\f{1}{2}\,\rho\,v_{i}\,v^{i}$ the total (internal plus kinetic) energy density, and $n_{e}$ the electron density.  
The pressure is related to the density, internal energy density, and electron density by an equation of state (EoS); i.e., $p=p(\rho,e,n_{e})$.  
(We also define the specific internal energy by $\epsilon=e/\rho$.)

@@

The mass conservation equation
\begin{equation}
  \pderiv{\rho}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\rho\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\rho\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\rho\,\bar{v}^{3}\,\Big)=0
\end{equation}
Components of the momentum equation
\begin{equation}
  \pderiv{\bar{M}^{i}}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\bar{P}^{1i}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\bar{P}^{2i}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\bar{P}^{3i}\,\Big)
  =\bar{S}^{i}+\bar{G}^{i},
\end{equation}
where the ``geometry sources'' are given by
\small
\begin{equation}
  \bar{S}^{1}
  =\bar{P}^{22}\f{1}{a}\pderiv{a}{x^{1}}+\bar{P}^{33}\f{1}{b}\pderiv{b}{x^{1}}, \quad
  \bar{S}^{2}
  =\bar{P}^{33}\f{1}{ac}\pderiv{c}{x^{2}}-\bar{P}^{12}\f{1}{a}\pderiv{a}{x^{1}}, \quad
  \bar{S}^{3}
  =-\bar{P}^{13}\f{1}{b}\pderiv{b}{x^{1}}-\bar{P}^{23}\f{1}{ac}\pderiv{c}{x^{2}},
\end{equation}
\normalsize
and the gravitational sources are
\begin{equation}
  \bar{G}^{1}
  =-\rho\,\pderiv{\Phi}{x^{1}}, \quad
  \bar{G}^{2}
  =-\rho\,\f{1}{a}\pderiv{\Phi}{x^{2}}, \quad
  \bar{G}^{3}
  =-\rho\,\f{1}{bc}\pderiv{\Phi}{x^{3}}.
\end{equation}
The energy equation is given by
\begin{equation}
  \pderiv{E}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\big(E+p\big)\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\big(E+p\big)\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\big(E+p\big)\,\bar{v}^{3}\,\Big)=Q,  
\end{equation}
where the gravitational energy source is
\begin{equation}
  Q=-\Big(\,\bar{M}^{1}\pderiv{\Phi}{x^{1}}+\bar{M}^{2}\f{1}{a}\pderiv{\Phi}{x^{2}}+\bar{M}^{3}\f{1}{bc}\pderiv{\Phi}{x^{3}}\,\Big).
\end{equation}
Finally, the electron conservation equation
\begin{equation}
  \pderiv{n_{e}}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,n_{e}\,\bar{v}^{1}\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,n_{e}\,\bar{v}^{2}\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,n_{e}\,\bar{v}^{3}\,\Big)=0
\end{equation}
By defining the vector of evolved variables $\bU=\big(\,\rho,\,\rho\,\bar{v}^{1},\,\rho\,\bar{v}^{2},\,\rho\,\bar{v}^{3},\,E,\,n_{e}\,\big)^{T}$, the vector of geometry sources $\vect{S}(\bU)=\big(\,0,\,\bar{S}^{1},\,\bar{S}^{2},\,\bar{S}^{3}\,,0,\,0\,\big)^{T}$, the vector of gravitational sources $\bG(\bU,\Phi)=\big(\,0,\bar{G}^{1},\bar{G}^{2},\bar{G}^{3},Q,0\,\big)^{T}$, and the vectors of flux vector components
\begin{equation}
  \vect{F}^{i}(\bU)
  =\big(\,\rho\,\bar{v}^{i},\,\rho\,\bar{v}^{1}\,\bar{v}^{i}+p\,\delta^{1i},\,\rho\,\bar{v}^{2}\,\bar{v}^{i}+p\,\delta^{2i},\,\rho\,\bar{v}^{3}\,\bar{v}^{i}+p\,\delta^{3i},\,(E+p)\,\bar{v}^{i},\,n_{e}\,\bar{v}^{i}\,\big)^{T},
\end{equation}
we can write the system of equations in compact form
\begin{equation}
  \pderiv{\bU}{t}
  +\f{1}{ab}\pderiv{}{x^{1}}\Big(\,ab\,\vect{F}^{1}(\bU)\,\Big)
  +\f{1}{ac}\pderiv{}{x^{2}}\Big(\,c\,\vect{F}^{2}(\bU)\,\Big)
  +\f{1}{bc}\pderiv{}{x^{3}}\Big(\,\vect{F}^{3}(\bU)\,\Big)
  =\vect{S}(\bU)+\bG(\bU,\Phi).
  \label{eq:extendedEulerCompact}
\end{equation}

@@

\bibliographystyle{apj}
\bibliography{../References/references.bib}
